Detection of Treatment-Subgroup Interactions in Clustered Datasets with Generalized Linear Mixed-effects Model Trees (GLMM trees)

### 2015-08-27 pkg/glmertree/paper/treatment_interactions.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@757 255f4249-2263-4417-b1ea-8498779e27a7

In the current paper, we present a tree-based algorithm for detecting treatment-subgroup interactions, which takes the clustered nature of datasets into account. The algorithm combines MOB with random-effects estimates, thus allowing for the detection of treatment-subgroup interactions, as well as accounting for variation between clusters (e.g., trials). In what follows, we will introduce the existing frameworks for estimating treatment effects: the GLM, model-based recursive partitioning, and the generalized linear mixed-effects model (GLMM). Then, we intoduce a new algorithm, which combines MOB and the GLMM: GLMM tree.

Before we discuss these methods for estimating treatment effects, we will introduce an artificial motivating data set, with which the methods will be illustrated. After we introduce the GLMM-tree algorithm, we present a simulation study, in which we evaluate GLMM trees comparative accuracy. Finally, in the application, we use GLMM tree to detect treatment-subgroup interactions in an existing dataset on the effects of treatments for depression.

@@

@@

@@

Where $y_i$ is the value of response variable for observation $i$, and $g$ is the link function, characterizing the relationship between the linear predictor $x_{i}^{\top}\beta$ and the mean of the response distribution function. In case of a continuous response variable, $g$ is often as taken as the identity function, and $\mu_{i}$ as the mean of a Gaussian distribution with variance $\sigma_{\epsilon}^2$. Further, $x_{i}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type (taking a value of 0 for the first, or reference treatment type, and a value of 1 for the second, or focal treatment type). $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, which is the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups.

@@

\vspace{20pt}
\begin{figure}[!htbp]
    \includegraphics[width=6cm]{glm_example.pdf}
	\caption{Example of a linear fixed-effects model for treatment outcomes, based on the artificial motivating dataset (N=150). The dot for Treatment 1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
	\label{fig:fixedeffects}
\end{figure}

@@

The rationale behind MOB is that a global model for all observations, like the GLM in Equation \ref{eq:expected_value} and \ref{eq:fixedeffects}, may not describe the data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}. This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different. However, CART trees detect differences in constant fits across terminal nodes, whereas MOB trees detect differences in parametric models across terminal nodes.

@@

\begin{figure}[!h]
    \includegraphics[width=15cm]{glmtree_example.pdf}
    \caption{Example of a tree representation of model-based recursive partition, based on the artificial motivating dataset. Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
    \label{fig:example_mobtree}
\end{figure}

@@

\begin{equation}
\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b_{m}
\end{equation}

Where $z_{i}$ is a vector of random-effects predictor variables values for observation $i$, and $b_{m}$ is the vector of random-effects regression coefficients in cluster $m$ ($m=1,...,M$), of which observation $i$ is part. Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in \citeA{BrykyRaud92}, for example.

For simplicity, we assume $z_{i}$ and $b_{m}$ to have length 1 in the current paper; that is, only cluster-specific intercepts are included in the models. However, random-effects covariates and coefficients can easily be included. Note that, alternatively, if the random-effects coefficients were known, values of $z_{i}^{\top}b_{m}$ could be included as an offset (i.e., a variable with a fixed coefficient of 1) in the GLM.

@@

\begin{equation}
\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b_{m}
\end{equation}

@@

\vspace{5pt}
\noindent Step 0: Initialize by setting $r$ and all values $\hat{b}_{m(r)}$ to 0.

\vspace{5pt}
\noindent \hangindent=4em Step 1: Set $r = r+1$. Estimate GLM tree ($x_{i}^{\top}\hat{\beta}_{j(r)}$) using responses $y_i - z_{i}^{\top}\hat{b}_{m(r-1)}$.

\vspace{5pt}
\noindent Step 2: Estimate random effects $z_{i}^{\top}\hat{b}_{m(r)}$ using responses $y_i - x_{i}^{\top}\hat{\beta}_{j(r)}$.

@@

\vspace{5pt}
The algorithm initializes by setting all values $b_m$ to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown. In every iteration, the GLM tree (i.e., the partition and corresponding fixed-effects coefficients $\beta_j$) and random-effects coefficients $b_m$ are re-estimated. The GLM tree is estimated, given the estimated $b_m$ values from the last iteration, and the $b_m$ values are estimated, given the estimated GLM tree from the current iteration. Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation \ref{eq:mixedeffects}.

@@

\begin{figure}[!h]
    \includegraphics[width=12cm]{glimmertree_example.pdf}
    \caption{Generalized Linear Mixed Model tree of the motivating example dataset. Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
    \label{fig:example_glimmertree}
\end{figure}

\section{Simulation}

@@

\begin{figure}[!h]
    \tikzstyle{decision} = [diamond, draw]
    \tikzstyle{line} = [draw, -stealth, thick]
    \tikzstyle{elli}=[draw, ellipse, minimum height=8mm, text width=6em, text centered]
   \tikzstyle{block} = [draw, rectangle, text width=7em, text centered, minimum height=15mm, node distance=6em]
    \begin{tikzpicture}[paths/.style={->, thick, >=stealth'}]
        \node [elli] (node1) { };
        \node [elli, below of =node1, xshift=-9em] (node2) { };
        \node [elli, below of =node1, xshift=9em] (node3) { };
        \node [block, below of=node2, xshift=-4em, yshift=-1em] (block1){\small$\beta_{j0}=17.5$\\$\beta_{j1}=-5.0$\\$d_j=-1.0$};
        \node [block, below of=node2, xshift=4em, yshift=-1em] (block2){\small$\beta_{j0}=30.0$\\$\beta_{j1}=0.0$\\$d_j=0.0$};
        \node [block, below of=node3, xshift=-4em, yshift=-1em] (block3){\small$\beta_{j0}=30.0$\\$\beta_{j1}=0.0$\\$d_j=0.0$};
        \node [block, below of=node3, xshift=5em, yshift=-1em, text width=8em] (block4){\small$\beta_{j0}=42.5$\\$\beta_{j1}=5.0$\\$d_j=1.0$};
        \draw [paths] (node1) to node [above left]{\small$U_2\leq30$} (node2);
        \draw [paths] (node1) to node [above right]{\small$U_2>30$} (node3);
        \draw [paths] (node2) to node [ left]{\small$U_1\leq17$} (block1);
        \draw [paths] (node2) to node [ right]{\small$U_1>17$} (block2);
        \draw [paths] (node3) to node [ left]{\small$U_5\leq63$} (block3);
        \draw [paths] (node3) to node [ right]{\small$U_5>63$} (block4);
    \end{tikzpicture}
	\caption{Data-generating model for treatment-subgroup interactions. Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment 1 and 2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
	\label{fig:modelC}
\end{figure}

@@

\begin{enumerate} 
\item Three levels for the total number of observations: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$, $K=15$ (where only $U_1$, $U_2$ and $U_5$ are true partitioning variables).
\item Two levels of intercorrelations between the covariates $U_1$ through $U_K$: $\rho_{U_{k},U_{k'}}=0.0$, $\rho_{U_{k},U_{k'}}=0.3$.
\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts are drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning covariate (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of about 0.42), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of about 0.42). 
\item Two different levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes, in subgroups with differential treatment effects. The levels for mean differences in subgroups with differential treatment effect were $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
\end{enumerate}

@@

\verb|R| \cite{R14} was used for generation and analysis of all datasets. Two additional R packages were used: \verb|partykit| (version 0.8-3; \citeNP{HothyZeil14,ZeilyHoth08}) for the estimation of GLM trees, and \verb|lme4| (version 1.1-7 \citeNP{BateyMeac12}) for the estimation of random-effects coefficients.

@@

@@

\begin{figure}[!htbp]
	\includegraphics[width=15cm]{xy_treesizes_maineff.pdf}
	\caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions. Values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=1$ represents the true tree size.}
	\label{fig:xyplot_treesize_nointeract}
\end{figure}

@@

\begin{figure}[!htbp]
	\includegraphics[width=15cm]{xy_treesizes_treatsubs.pdf}
	\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions. Values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=7$ represents true tree size.}
	\label{fig:xyplot_treesize_interact}
\end{figure}

@@

\begin{figure}[!htbp]
	\includegraphics[width=15cm]{xy_accuracy.pdf}
	\caption{Average accuracy of GLM and GLMM trees. Accuracy of trees is defined as the proportion of datasets in which the true tree was accurately recovered. Values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size.}
	\label{fig:xyplot_treeaccuracy}
\end{figure}

@@

\begin{figure}[!htbp]
	\includegraphics[width=12cm]{xy_correlations.pdf}
	\caption{Average predictive accuracy of GLM and GLMM trees. Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment 1 and 2. Values $5$ and $2.5$ refer to the absolute value of the unstandardized treatment-effect difference in subgroups with treatment-effect differences; values $200$, $500$ and $1000$ refer to sample size.}
	\label{fig:xyplot_correlations}
\end{figure}

\section{Application to real data}

@@

\begin{figure}[!ht]
    \includegraphics[width=12cm]{app_lmtree.pdf}
	\caption{GLM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs. pharmacotherapy (PHA).}
	\label{fig:lmtree_C&W}
\end{figure}

@@

In the current paper, we presented the GLMM-tree algorithm, which allows for the estimation of a GLM-based recursive partition, as well as the estimation of random-effects parameters. The results of our simulation study show that GLMM tree performed very well in recovering treatment-subgroup interactions, by recovering the true tree structure in 90\% of the simulated datasets with treatment-subgroup interactions. In contrast, GLM tree recovered the true tree structure in only 61\% of the datasets with treatment-subgroup interactions. In the absence of treatment-subgroup interactions, GLMM tree erroneously detected subgroups in only 4\% of the datasets, whereas GLM tree erroneously detected subgroups in 33\% of those datasets.

@@

Although these findings are encouraging for the use of GLMM tree in the detection of treatment-subgroup interactions in datasets with clustered structures, some limitations and challenges for future research should be noted.

@@

A challenge for future research is the development of more adequate ways to deal with missing data. GLM tree, and therefore also GLMM tree, handle missing data by listwise deletion, like most tree-based algorithms for treatment-subgroup interaction detection. However, missing data commonly occurs in clinical trails, and listwise deletion is not an optimal approach for dealing with missing data.

@@

\newpage
\section{Appendix: Notation}

>>>>

In the current paper, we present a tree-based algorithm for detecting treatment-subgroup interactions, which takes the clustered nature of datasets into account. The algorithm combines MOB with random-effects estimates, thus allowing for the detection of treatment-subgroup interactions, as well as accounting for variation between clusters (e.g., trials). In what follows, we will introduce the existing frameworks for estimating treatment effects: the GLM, model-based recursive partitioning, and the generalized linear mixed-effects model (GLMM). Then, we intoduce a new algorithm, which combines MOB and the GLMM: GLMM tree. Subsequently, we evaluate the performances of the GLMM-tree algorithm in simulated datasets, and illustrate application of the algorithm with an exisiting dataset of a patient-level meta-analysis on the effects of treatments for depression by \citeA{CuijyWeit14}.

Before we discuss the methods for estimating treatment effects, we will introduce an artificial motivating data set, with which the methods will be illustrated. After we introduce the GLMM-tree algorithm, we present a simulation study, in which we evaluate GLMM trees comparative accuracy. Finally, in the application, we use GLMM tree to detect treatment-subgroup interactions in an existing dataset on the effects of treatments for depression.

@@

<<artificial-data>>=
set.seed(123)
treatment <- rbinom(n=150,size=1,prob=1/2)
duration <- round(matrix(rnorm(150,mean=5,sd=3),ncol=1))
duration[duration<0] <- 1
anxiety <- round(rnorm(150,mean=10,sd=3))
age <- round(rnorm(150,mean=30,sd=10))
age[age<18] <- 18
age[age>75] <- 75
cutscore1 <- 5+.44*3
cutscore2 <- 10
node1t1 <- ifelse(duration<=cutscore1 & anxiety<=cutscore2 & treatment==0, 1, 0)
node1t2 <- ifelse(duration<=cutscore1 & anxiety<=cutscore2 & treatment==1, 1, 0)
node2t1 <- ifelse(duration<=cutscore1 & anxiety>cutscore2 & treatment==0, 1, 0)
node2t2 <- ifelse(duration<=cutscore1 & anxiety>cutscore2 & treatment==1, 1, 0)
node3t1 <- ifelse(duration>cutscore1 & treatment==0, 1, 0)
node3t2 <- ifelse(duration>cutscore1 & treatment==1, 1, 0)
beta1 <- -2
beta2 <- 0
beta3 <- 2.5
error <- rnorm(150,0,2)
cluster <- error+rnorm(150, 0, 6)
int_values <- rnorm(10, 0, 1)
tmp <- rep(int_values,each=15)
tmp <- tmp[order(tmp)]
rand_int <- rep(NA, times=150)
rand_int[order(cluster)] <- tmp
cluster[order(cluster)] <- rep(1:10, each=15)
cor(error,rand_int)
true_error <- error - rand_int
true_eff <- round(9 + beta1*node1t1 - beta1*node1t2 + beta2*node2t1 - beta2*node2t2 + beta3*node3t1 - beta3*node3t2 + 
     + .4*treatment)
outcome <- round(9 + beta1*node1t1 - beta1*node1t2 + beta2*node2t1 - beta2*node2t2 + beta3*node3t1 - beta3*node3t2 + 
                      + .4*treatment + true_error + rand_int)
outcome[outcome<0] <- 1
example <- data.frame(treatment,outcome,duration,anxiety,cluster, true_eff)
example$treatment <- factor(example$treatment,labels=c("Treatment 1","Treatment 2"))
#cor(outcome,rand_int)^2
#cor(outcome, true_eff)^2
#truetree <- lmtree(true_eff ~ treatment | age + duration + anxiety, data=example)
@

@@

<<example-trees>>=
exampleglm <- lmtree(outcome ~ treatment | age, data=example)
exampleglmtree <- lmtree(outcome ~ treatment | age + duration + anxiety, data=example)
exampleglmmtree <- lmertree(outcome ~ treatment | age + duration + anxiety, outcome ~ (1|cluster), data=example)
@

@@

Where $y_i$ is the value of response variable for observation $i$, and $g$ is the link function, characterizing the relationship between the linear predictor $x_{i}^{\top}\beta$ and the mean of the response distribution function. In case of a continuous response variable, $g$ is often as taken as the identity function, and $\mu_{i}$ as the mean of a Gaussian distribution with variance $\sigma_{\epsilon}^2$. Further, $x_{i}$ is a vector of fixed-effects predictor variable values for observation $i$, of which the first element takes a value of 1 for the intercept, and the second element takes the value of a dummy indicator for treatment type (a value of 0 for the first, or reference treatment type, and a value of 1 for the second, or focal treatment type). $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, which is the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups.

@@

\begin{figure}[!ht]
\setkeys{Gin}{width=0.5\textwidth}
<<exampleglm, fig=TRUE, height=5, width=5>>=
plot(exampleglm)
@
	\caption{Example of a linear fixed-effects model for treatment outcomes, based on the artificial motivating dataset (N=150). The dot for Treatment 1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
	\label{fig:fixedeffects}
\end{figure}

@@

The rationale behind MOB is that a global model for all observations, like the GLM in Equation \ref{eq:expected_value} and \ref{eq:fixedeffects}, may not describe all data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}. This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different. However, CART trees detect differences in constant fits across terminal nodes, whereas MOB trees detect differences in parametric models across terminal nodes.

@@

\begin{figure}[!h]
\setkeys{Gin}{width=\textwidth}
<<exampleglmtree, fig=TRUE, width=12, height=9>>=
plot(exampleglmtree)
@
    \caption{Example of a tree representation of model-based recursive partition, based on the artificial motivating dataset. Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
    \label{fig:example_mobtree}
\end{figure}

@@

\begin{equation}
\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}

Where $z_{i}$ is a unit vector of length $M$, of which the $m$th element takes a value of 1, and all other elements take a value of 0, with $m$ ($m=1,...,M$) denotes the cluster of which observation $i$ is part. $b$ is random vector of length $M$, each element being the random intercept for $m$. Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in \citeA{BrykyRaud92}, for example.

For simplicity, we assume that only cluster-specific intercepts are included in the models. However, random-effects covariates and coefficients can easily be included. Note that, alternatively, if the random-effects coefficients were known, values of $z_{i}^{\top}b$ could be included as an offset (i.e., a variable with a fixed coefficient of 1) in the GLM.

@@

\begin{equation}
	\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
\end{equation}

@@

\vspace{5pt}
\noindent Step 0: Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.

\vspace{5pt}
\noindent \hangindent=4em Step 1: Set $r = r+1$. Estimate GLM tree ($x_{i}^{\top}\hat{\beta}_{j(r)}$) using $z_{i}^{\top}\hat{b}_{(r-1)}$ as an offset.

\vspace{5pt}
\noindent Step 2: Estimate random effects $z_{i}^{\top}\hat{b}_{(r)}$ using $x_{i}^{\top}\hat{\beta}_{j(r)}$ as an offset.

@@

\vspace{5pt}
The algorithm initializes by setting all $b$ values to $0$, since the random-effects (and also the fixed-effects) parts are initially unknown. In every iteration, the GLM tree (i.e., the partition and corresponding fixed-effects coefficients $\beta_j$) and random-effects coefficients $b_m$ are re-estimated. The GLM tree is estimated, given the estimated $\hat{b}$ from the last iteration, and the $b$ values are estimated, given the estimated GLM tree from the current iteration. Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation \ref{eq:mixedeffects}.

\begin{figure}[!h]
<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
plot(exampleglmmtree$Tree)
@
    \caption{Generalized Linear Mixed Model tree of the motivating example dataset. Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
    \label{fig:example_glimmertree}
\end{figure}

@@

\section{Emperical evaluation}

@@

\begin{figure}[!h]
\setkeys{Gin}{width=0.8\textwidth}
<<dgp-tree, fig=TRUE, height=5, width=7>>=
fig4 <- party(
partynode(1L,
 split = partysplit(2L, breaks = 30),
 kids = list(
   partynode(2L,
     split = partysplit(1L, breaks = 17),
     kids = list(
       partynode(3L, info = c(
  expression(''),
  expression(beta[j0] == '17.5'),
  expression(''),
  expression(beta[j1] == -'5.0'),
  expression(''),
  expression(d[j] == -'1.0')
 )),
       partynode(4L, info = c(
  expression(''),
  expression(beta[j0] == '30.0'),
  expression(''),
  expression(beta[j1] == '0.0'),
  expression(''),
  expression(d[j] == '0.0')
 )))),
   partynode(5L,
     split = partysplit(1L, breaks = 63),
     kids = list(
       partynode(6L, info = c(
  expression(''),
  expression(beta[j0] == '30.0'),
  expression(''),
  expression(beta[j1] == '0.0'),
  expression(''),
  expression(d[j] == '0.0')
 )),
       partynode(7L, info = c(
  expression(''),
  expression(beta[j0] == '42.5'),
  expression(''),
  expression(beta[j1] == '5.0'),
  expression(''),
  expression(d[j] == '1.0')
 )))))),
data.frame(U1 = numeric(0), U2 = numeric(0))
)
plot(fig4, tnex=1.5)
@
\caption{Data-generating model for treatment-subgroup interactions. Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment 1 and 2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
\label{fig:modelC}
\end{figure}

@@

\begin{enumerate} 
\item Three levels for the total number of observations: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$, $K=15$ (where only $U_1$, $U_2$ and $U_5$ are true partitioning variables).
\item Two levels of intercorrelations between the covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts are drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning covariate (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of about 0.42), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of about 0.42). 
\item Two different levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes, in subgroups with differential treatment effects. The levels for mean differences in subgroups with differential treatment effect were $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
\end{enumerate}

@@

\verb|R| \cite{R14} was used for generation and analysis of all datasets. Two additional R packages were used: \verb|partykit| (version 1.0-2; \citeNP{HothyZeil14,ZeilyHoth08}) for the estimation of GLM trees, and \verb|lme4| (version 1.1-7 \citeNP{BateyMeac12}) for the estimation of random-effects coefficients.

@@

<<simulation-data>>=
load("treespecs_long.dat")
treespecs.long$N <- factor(as.numeric(as.character(treespecs.long$N)), ordered=T)
treespecs.long$sigmab <- factor(as.numeric(as.character(treespecs.long$sigmab)), ordered=T)
treespecs.long$treatdiff <- factor(as.numeric(as.character(treespecs.long$treatdiff)), ordered=T)
treespecs.long$truetree.values <- as.numeric(treespecs.long$truetree.values)-1
levels(treespecs.long$corUb)[levels(treespecs.long$corUb)=="bi and splitting U correlated"] <- "b correlated with splitting U"
levels(treespecs.long$corUb)[levels(treespecs.long$corUb)=="bi and non-splitting U correlated"] <- "b correlated with non-splitting U"
levels(treespecs.long$corUb)[levels(treespecs.long$corUb)=="uncorrelated"] <- "b and U uncorrelated"
@

@@

\begin{figure}[!htbp]
<<treesize-nointeract-xyplot, fig=TRUE, width=8, height=7>>=
load("treesizes_long_nointeract.dat")
treesizes.long$sigmabm <- factor(as.numeric(as.character(treesizes.long$sigmabm)), ordered=T)
treesizes.long$N <- factor(as.numeric(as.character(treesizes.long$N)), ordered=T)
aggdata <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmabm + corUbm, FUN=mean, data=treesizes.long)
print(xyplot(treesize.values ~ sigmabm | N + corUbm, data = aggdata, groups=treesize.ind, type="b", ylab="tree size", xlab="sigma_b", par.settings=standard.theme("pdf",color=F), abline=c(1,0), auto.key=list(space="top", columns=2, title="Algorithm type", cex.title=1,lines=T, points=T)))
@
\caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions. Values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=1$ represents the true tree size.}
\label{fig:xyplot_treesize_nointeract}
\end{figure}

@@

\begin{figure}[!htbp]
<<treesize-xyplot, fig=TRUE, width=9, height=9.5>>=
aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)	
print(xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups=treesize.ind, type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), abline=c(7,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T)))
@
\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions. Values $200$, $500$ and $1000$ refer to sample size. Reference line at $y=7$ represents true tree size.}
\label{fig:xyplot_treesize_interact}
\end{figure}

@@

\begin{figure}[!htbp]
<<accuracy-xyplot, fig=TRUE, width=9, height=9.5>>=
aggdata.acc <- aggregate(formula=truetree.values ~ truetree.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)
print(xyplot(truetree.values ~ sigmab | N + corUb, data = aggdata.acc, groups=truetree.ind, type="b",   ylab="tree accuracy", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T)))
@
\caption{Average accuracy of GLM and GLMM trees. Accuracy of trees is defined as the proportion of datasets in which the true tree was accurately recovered. Values are correlated to one of the $U_k$ variables; values $200$, $500$ and $1000$ refer to sample size.}
\label{fig:xyplot_treeaccuracy}
\end{figure}

@@

\begin{figure}[!htbp]
<<correlation-xyplot, fig=TRUE>>=
aggdata.cor <- aggregate(formula=correlation.values ~ correlation.ind + N + sigmab + treatdiff, FUN=mean, data=treespecs.long)
print(xyplot(correlation.values ~ sigmab | N + treatdiff, data = aggdata.cor, groups=correlation.ind, type="b", ylab="correlation", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T)))
@
\caption{Average predictive accuracy of GLM and GLMM trees. Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment 1 and 2. Values $5$ and $2.5$ refer to the absolute value of the unstandardized treatment-effect difference in subgroups with treatment-effect differences; values $200$, $500$ and $1000$ refer to sample size.}
\label{fig:xyplot_correlations}
\end{figure}

\section{Application to patient-level meta-analysis dataset on the effects of treatments for depression}

<<depression-data>>=
metadata <- read.dta("Database IPDMA CBT PHA Version 11.dta")
metadata[metadata == 999] <- NA
metadata[metadata == 888] <- NA
vars <- c("studyid", "Tx_group", "Age", "Gender", "education", "ComorbidAnxietyDisorder", "HRSDt0", "HRSDt1")
factors <- c("studyid", "Tx_group", "Gender", "education", "ComorbidAnxietyDisorder")
metadata$education <- factor(metadata$education, ordered = TRUE)
for (i in 1:length(factors)) {metadata[, factors[i]] <- factor(metadata[,factors[i]])}
metadata <- metadata[vars] # select only relevant variables
metadata <- metadata[complete.cases(metadata[, vars]), ] # select only complete data
metadata <- metadata[!metadata$Tx_group == "placebo", ] # remove placebo observations
metadata$Tx_group <- factor(metadata$Tx_group)
@

@@

\begin{figure}[!ht]
\setkeys{Gin}{width=0.75\textwidth}
<<depression-glmtree, fig=TRUE, height=6, width=8>>=
plot(lmtree_app)
@
\caption{GLM tree for prediction of posttreatment total scores on the Hamilton Rating Scale for Depression (HAM-D). The y-axes of the boxplots represent posttreatment HAM-D scores, and the x-axes represent treatment levels: cognitive behavior therapy (CBT) vs. pharmacotherapy (PHA).}
\label{fig:lmtree_C&W}
\end{figure}

@@

In the current paper, we presented the GLMM-tree algorithm, which allows for the estimation of a GLM-based recursive partition, as well as the estimation of random-effects parameters. The results of our simulation study show that GLMM tree performed very well in recovering treatment-subgroup interactions, by recovering the true tree structure in 90\% of the simulated datasets with treatment-subgroup interactions. In contrast, GLM tree recovered the true tree structure in only 61\% of the datasets with treatment-subgroup interactions. In the absence of treatment-subgroup interactions, GLMM tree erroneously detected subgroups in only 4\% of the datasets, whereas GLM tree erroneously detected subgroups in 33\% of those datasets. In other words, the Type I error rate of GLMM tree very closely resembled the $\alpha$ level used for evaluating significance of parameter instability, whereas the Type I error rate of GLM tree clearly exceeded this value.

@@

Although these findings are encouraging for the use of GLMM tree in the detection of treatment-subgroup interactions in datasets with clustered structures, some limitations should be noted.

@@

@@

\section{Appendix: Notation}
